How quantum mechanics requires non-additive measures

### 2023-10-28 Proceedings/2023-QIP/QuantumMeasure.tex
Fix abstract and fix/add references

\begin{abstract}
	Measure theory is used in physics, not just to capture classical probability, but also to quantify the number of states. In previous works, in fact, we found that state quantification plays a foundational role in classical mechanics, and, therefore, we set ourselves to construct the quantum equivalent of the Liouville measure. Unlike the classical counterpart, this quantized measure is non-additive and has a unitary lower bound (i.e. no set of states cannot have less than one state). Conversely, requiring that state quantification is finite for finite continuous regions and unitary for each state already implies non-additivity, which implies the failure of classical theory. In this article we show these preliminary results and outline a new line of inquiry that may provide a different insights into the foundations of quantum theory. Additionally, this new approach may turn out to be useful to those interested in a quantized theory of space-time, as we believe this requires a quantized measure for the quantification of the independent degrees of freedom.
\end{abstract}

@@

It is well-established that standard measure theoretic tools cannot be used to characterize quantum probability as quantum theory is a non-classical (i.e. non-Kolmogorov) probability theory. To obviate this problem, various alternatives and extensions to measure theory and probability theory have been developed with different goals.\cite{groenewold1946principles, moyal1949quantum, gleason1975measures, sorkin1994quantum, hamhalter2013quantum, svozil2022extending, monchietti2023measure} Our interest in measure theory, however, is not in its use to capture classical probability. Measures play another crucial use in physical theories: ``counting'', or better ``quantifying'', states. It is this use that will be the focus of this work.

As part of our overall project Assumptions of Physics,\cite{aop-book} we have developed an approach, called reverse physics,\cite{aop-phys-ReversePhysics} which, starting from the laws, aims to find a minimal set of physical assumptions needed to rederive them. We found that the ability to quantify states is a crucial component of classical theory, so much so that it is enough to recover the structure of phase space and the laws of evolution.\cite{aop-phil-HamiltonianPrivilege,aop-phil-Hamiltonianinformation} The former (i.e. the symplectic structure) is the only structure that allows one to quantify states and configurations over independent degrees of freedom in a way that is frame independent. If one imposes deterministic and reversible evolution, state quantification must be preserved, leading to Hamiltonian evolution (i.e. a symplectomorphism). All elements of classical mechanics, including canonical variables, the Lagrangian and the action principle, \cite{aop-phys-ActionPrincipleInterpretation} can be rederived and understood in terms of quantifying the flow of states. Given the formal similarity between classical and quantum mechanics, we believe a similar approach is possible for the latter. This raises the question: what is the quantum analogue of the Liouville measure, the measure that quantifies states in the classical theory?

@@

The aim of this article is to propose and articulate the goals of this new line of research. It provides the overall picture in which future works will add further mathematical detail and physical insights. In section II we will show how measure theory is used to quantify elements in a set and how this is needed to assign probability and probability densities to each element. In section III we will see how statistical mechanics and information theory mandate a link between entropy and state quantification, which we will use to define the quantum analogue of the Liouville measure. We will find that this measure is not additive.\footnote{Technically, measures are defined to be additive, therefore what we find is not a measure, but a set function. However, this term is not widespread, so we use the oxymoron ``non-additive measure'' as it is customary in the mathematical literature. \cite{nonadditive2016, denneberg1994non}} In section IV we will see that this non-additivity is linked to non-contextuality. In section V we will see that quantization means putting a unitary lower bound on the quantification of states, while keeping a finite value over finite regions. We will see that this very directly requires non-additivity and therefore non-contextuality. In section VI we will argue that quantization of space-time means using a quantized measure on the quantification of degrees of freedom. Lastly, in section VII we will outline a series of goals for a research program that aims to fully characterize the quantized measure.

@@

However, the structure of classical phase space (i.e. symplectic structure) is exactly the structure that allows one to define a quantifying measure for states that is invariant under change of frame.\cite{aop-book,aop-phil-HamiltonianPrivilege,aop-phil-Hamiltonianinformation} The role of the quantifying measure in classical mechanics, then, is assigned to the Liouville measure, which we can understand as a Lebesgue measure on position and conjugate momentum (i.e. canonical coordinates).\footnote{The volume of an infinitesimal parallelepiped is given by $dq^1 dq^2 \cdots dp_1 dp_2 \cdots$. Since $dq^i$ is contravariant and $dp_i$ is covariant under change of frame, the volume remains the same.} As we shrink the size of the set, $\frac{p(U)}{\mu(U)}$ becomes the Radon-Nikodym derivative $\frac{dp}{d\mu}$, which gives us the probability density, not the probability itself.

@@

So, in light of this, what is quantization? If we have a system characterized by continuous quantities, such as position, the state space will necessarily be a continuous space and finite regions will have to correspond to finite information as those correspond to the finite precision measurements we can make.\footnote{For a treatment of how finite-time experimental verification is captured by topological concepts, and what necessary and sufficient operational assumptions are required for the use of real valued quantities, see \cite{aop-book}} If the regions are large enough, the Liouville (or Lebesgue) measure will work. But as we shrink a region we will encounter, given the additivity, regions that contain less than one state, which does not make sense. Note that measure less than one corresponds to negative thermodynamic entropy, which does not make sense either.

@@

While we have found some literature on non-additive measures and set functions \cite{hamhalter2013quantum, nonadditive2016}, together with modified theories of integration, these seem to concentrate on monotonic measures: measures that increase as the set increases. Unfortunately, the quantized measure does not have this feature. Consider the set $\{ |z^+\>, |z^-\> \}$. An equal mixture will return the maximally mixed state which, as we said, would correspond to a measure equal to 2. Now consider the set $\{ |z^+\>, |z^-\>, |x^+\> \}$. An equal mixture of them will not correspond to the maximally mixed state, and therefore the measure will be less than two. The set is bigger, but the measure is smaller. Therefore there may be new mathematical ground to be explored.

>>>>

\begin{abstract}
	Measure theory is used in physics, not just to capture classical probability, but also to quantify the number of states. In previous works, we found that state quantification plays a foundational role in classical mechanics, and therefore, we set ourselves to construct the quantum equivalent of the Liouville measure. Unlike the classical counterpart, this quantized measure is non-additive and has a unitary lower bound (i.e. no set of states can have less than one state). Conversely, requiring that state quantification is finite for finite continuous regions and that each state counts as one already implies non-additivity, which in turn implies the failure of classical theory. In this article we show these preliminary results and outline a new line of inquiry that may provide a different insight into the foundations of quantum theory. Additionally, this new approach may prove to be useful to those interested in a quantized theory of space-time, as we believe this requires a quantized measure for the quantification of the independent degrees of freedom.
\end{abstract}

@@

It is well-established that standard measure theoretic tools cannot be used to characterize quantum probability as quantum theory is a non-classical (i.e. non-Kolmogorov) probability theory. To obviate this problem, various alternatives and extensions to measure theory and probability theory have been developed with different goals.\cite{groenewold1946principles, moyal1949quantum, gleason1957measures, sorkin1994quantum, hamhalter2003quantum, gudder2009quantum, svozil2022extending, monchietti2023measure} Our interest in measure theory, however, is not in its use to capture classical probability. Measures play another crucial use in physical theories: ``counting'', or better ``quantifying'', states. It is this use that will be the focus of this work.

As part of our overall project Assumptions of Physics,\cite{aop-book} we have developed an approach, called reverse physics,\cite{aop-phys-ReversePhysics} which, starting from the laws, aims to find a minimal set of physical assumptions needed to rederive them. We found that the ability to quantify states is a crucial component of classical theory, so much so that it is enough to recover the structure of phase space and the laws of evolution.\cite{aop-phil-Hamiltonianinformation,aop-phil-HamiltonianPrivilege} The former (i.e. the symplectic structure) is the only structure that allows one to quantify states and configurations over independent degrees of freedom in a way that is frame independent. If one imposes deterministic and reversible evolution, state quantification must be preserved, leading to Hamiltonian evolution (i.e. a symplectomorphism). All elements of classical mechanics, including canonical variables, the Lagrangian and the action principle, \cite{aop-phys-ActionPrincipleInterpretation} can be rederived and understood in terms of quantifying the flow of states. Given the formal similarity between classical and quantum mechanics, we believe a similar approach is possible for the latter. This raises the question: what is the quantum analogue of the Liouville measure, the measure that quantifies states in the classical theory?

@@

The aim of this article is to propose and articulate the goals of this new line of research. It provides the overall picture in which future works will add further mathematical detail and physical insights. In section II we will show how measure theory is used to quantify elements in a set and how this is needed to assign probability and probability densities to each element. In section III we will see how statistical mechanics and information theory mandate a link between entropy and state quantification, which we will use to define the quantum analogue of the Liouville measure. We will find that this measure is not additive.\footnote{Technically, measures are defined to be additive, therefore what we find is not a measure, but a set function. However, this term is not widespread, so we use the oxymoron ``non-additive measure'' as it is customary in the mathematical literature. \cite{denneberg1994non,nonadditive2014}} In section IV we will see that this non-additivity is linked to non-contextuality. In section V we will see that quantization means putting a unitary lower bound on the quantification of states, while keeping a finite value over finite regions. We will see that this very directly requires non-additivity and therefore non-contextuality. In section VI we will argue that quantization of space-time means using a quantized measure on the quantification of degrees of freedom. Lastly, in section VII we will outline a series of goals for a research program that aims to fully characterize the quantized measure.

@@

However, the structure of classical phase space (i.e. symplectic structure) is exactly the structure that allows one to define a quantifying measure for states that is invariant under change of frame.\cite{aop-book,aop-phil-Hamiltonianinformation,aop-phil-HamiltonianPrivilege} The role of the quantifying measure in classical mechanics, then, is assigned to the Liouville measure, which we can understand as a Lebesgue measure on position and conjugate momentum (i.e. canonical coordinates).\footnote{The volume of an infinitesimal parallelepiped is given by $dq^1 dq^2 \cdots dp_1 dp_2 \cdots$. Since $dq^i$ is contravariant and $dp_i$ is covariant under change of frame, the volume remains the same.} As we shrink the size of the set, $\frac{p(U)}{\mu(U)}$ becomes the Radon-Nikodym derivative $\frac{dp}{d\mu}$, which gives us the probability density, not the probability itself.

@@

So, in light of this, what is quantization? If we have a system characterized by continuous quantities, such as position, the state space will necessarily be a continuous space and finite regions will have to correspond to finite information as those correspond to the finite precision measurements we can make.\footnote{For a treatment of how finite-time experimental verification is captured by topological concepts, and what necessary and sufficient operational assumptions are required for the use of real valued quantities, see \cite{kelly1996,aop-book}.} If the regions are large enough, the Liouville (or Lebesgue) measure will work. But as we shrink a region we will encounter, given the additivity, regions that contain less than one state, which does not make sense. Note that measure less than one corresponds to negative thermodynamic entropy, which does not make sense either.

@@

While we have found some literature on non-additive measures and set functions \cite{hamhalter2003quantum, nonadditive2014}, together with modified theories of integration, these seem to concentrate on monotonic measures: measures that increase as the set increases. Unfortunately, the quantized measure does not have this feature. Consider the set $\{ |z^+\>, |z^-\> \}$. An equal mixture will return the maximally mixed state which, as we said, would correspond to a measure equal to 2. Now consider the set $\{ |z^+\>, |z^-\>, |x^+\> \}$. An equal mixture of them will not correspond to the maximally mixed state, and therefore the measure will be less than two. The set is bigger, but the measure is smaller. Therefore there may be new mathematical ground to be explored.
